---
title: "Time Mixed Effects Model"
author: 
  -"Aissata Bah"
  -"Brice Laurent"
  -"Linh Vu"
date: "2023-12-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r,message=F}
# Load Libraries
library(dplyr)
library(ggplot2)
library(lme4)
library(broom.mixed)
library(lmerTest)

data = read.csv("./data/data_clean.csv")
```

# Theory Behind Using Linear Mixed Effects Model for Time

$$HDI_{year,region} = \alpha_{region}+\beta_{region}X_{year,country,region}+\varepsilon_{country,region}$$
where $\alpha_{region}\sim N(\mu_{\alpha},\sigma_{\alpha}^2)$ and $\beta_{region}\sim N(\mu_{\beta},\sigma_{\beta}^2)$

This equation is assuming no correlation between the random effects. However, from the following plot, we can see that the average HDI of all the regions seem to follow a similar trajectory over time and, therefore, we would expect the random effects to be correlated between regions. Because of this, we will allow for correlation between random effects, for which there is a closed form solution. 

```{r}
average_hdi = data %>% group_by(region,year) %>% summarise(average_hdi = mean(hdi))
average_hdi[average_hdi$year == "2012-2013", "year"] = "2012.5"
average_hdi[average_hdi$year == "2017-2018", "year"] = "2017.5"
ggplot(data=average_hdi, aes(x=as.numeric(year), y=average_hdi)) + geom_line(aes(color = region))
```


```{r}
# Create Time Mixed Effects Model
selected_columns = data[, c("country","year", "region", "hdi", "x1.6","x3.2", "x5.1", "x6.4", "x7.3")]
selected_columns[selected_columns$year == "2012-2013", "year"] = "2012.5"
selected_columns[selected_columns$year == "2017-2018", "year"] = "2017.5"
selected_columns$year = as.numeric(selected_columns$year)
selected_columns$years_from_2000 = selected_columns$year - 2000

time_mixed_effects = lmer(hdi~x1.6+x3.2+x5.1+x6.4+x7.3+(1+x1.6+x3.2+x5.1+x6.4+x7.3|region)+(1+years_from_2000||country),data=selected_columns)
```

```{r}
# Determine Significance of Coefficients
fixed_effects = summary(time_mixed_effects)$coefficients
fixed_effects
```
Degrees of Freedom=Total number of observations−Number of fixed effects−Number of estimated variance components

Bonferroni Correction: $\alpha$=0.05/6=0.00833

```{r}
# Importance of Considering Region
time_mixed_effects_noRegion = lmer(hdi~x1.6+x3.2+x5.1+x6.4+x7.3+(1+years_from_2000||country),data=selected_columns)
anova(time_mixed_effects,time_mixed_effects_noRegion)

# Importance of Considering Year
time_mixed_effects_noYear = lmer(hdi~x1.6+x3.2+x5.1+x6.4+x7.3+(1+x1.6+x3.2+x5.1+x6.4+x7.3|region),data=selected_columns)
anova(time_mixed_effects,time_mixed_effects_noYear)
```

```{r}
# Make Plots
country_coef = as.data.frame(coef(time_mixed_effects)$country)
country_coef = merge(country_coef, selected_columns[,c("country", "region")], by.x=c("row.names"), by.y=c("country"))

ggplot(data = country_coef, aes(x=`(Intercept)`, y = years_from_2000)) + geom_point(aes(color=region))

residuals = augment(time_mixed_effects)
ggplot(data = residuals, aes(x=.fitted, y = .resid)) +
  geom_point(aes(color = region)) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  geom_hline(yintercept = 0) +
  labs(y="Residuals", x="FItted HDI")
```




















```{r}
# Old wrong model
time_mixed_effects = lmer(hdi~years_from_2000+x1.6+x3.2+x5.1+x6.4+x7.3+(1+years_from_2000|region),data=selected_columns)


residuals = augment(time_mixed_effects)
ggplot(data = residuals, aes(x=.fitted, y = .resid)) +
  geom_point(aes(color = region)) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  geom_hline(yintercept = 0) +
  labs(y="Residuals", x="Fitted HDI")
```