---
title: "TimeMixedEffectsModel"
author: 
  -"Aissata Bah"
  -"Brice Laurent"
  -"Linh Vu"
date: "2023-12-07"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
```

```{r}
# Load Libraries
library(dplyr)
library(ggplot2)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(ggpubr)

data = read.csv("./data/data_clean.csv")
```

# Mixed Effects Linear Model for Time

To include the data points from all the years collected in both datasets, we decided to incorporate time effects into a mixed effects model. For this model, we  used the variables manually selected based on correlations: `x1.6`, `x3.2`, `x5.1`, `x6.4`, and `x7.3`. This mixed effects model is based on the equation:
$$HDI_{time,~country,~region} = \alpha_{region}+\beta_{region}\times X_{time,~country,~region}+a_{country}+b\times X_{time,~country}+\varepsilon_{time,~country,~region}$$
where $\alpha_{region}\sim N(\mu_{\alpha},\sigma_{\alpha}^2)$, $\beta_{region}\sim N(\mu_{\beta},\sigma_{\beta}^2)$, $a_{country}\sim N(\mu_{a},\sigma_{a}^2)$, $b_{country}\sim N(\mu_{b},\sigma_{b}^2)$, and $\varepsilon_{time,~country,~region}\sim N(0,\sigma_{HDI}^2)$.

$X_{time,~country,~region}$ is the design matrix for the variables `x1.6`, `x3.2`, `x5.1`, `x6.4`, `x7.3`, and intercept, while $X_{time,~country}$ is the design matrix for the time variable (in years) and intercept. 

This equation above assumes independence between random effects. Though this assumption is reasonable for the random effects of Rule of Law variables between regions, it is not as reasonable for the random effects of the time variable between countries. This can be seen from the chart below since the HDI of different countries mostly follow similar trajectories over times, and therefore are associated. There is no closed form solution when we allow correlation between random effects, so the actual mixed effects model will not exactly follow the equation above.

```{r}
data[data$year == "2012-2013", "year"] = "2012.5"
data[data$year == "2017-2018", "year"] = "2017.5"
data$year = as.numeric(data$year)
average_hdi = data %>% group_by(region,year) %>% summarise(average_hdi = mean(hdi))
hdi_region = ggplot(data=average_hdi, aes(x=year, y=average_hdi)) + geom_line(aes(color = region))+  theme(legend.text = element_text(size = 8))
hdi_country = ggplot(data=data, aes(x=year, y=hdi, group = country, col=region)) + geom_line()+  theme(legend.text = element_text(size = 8))
ggarrange(hdi_region, hdi_country, ncol=2, common.legend = TRUE, legend="bottom")
```


```{r}
# Create Time Mixed Effects Model
selected_columns = data[, c("country","year", "region", "hdi", "x1.6","x3.2", "x5.1", "x6.4", "x7.3")]
selected_columns[selected_columns$year == "2012-2013", "year"] = "2012.5"
selected_columns[selected_columns$year == "2017-2018", "year"] = "2017.5"
selected_columns$year = as.numeric(selected_columns$year)
selected_columns$years_from_2000 = selected_columns$year - 2000

time_mixed_effects = lmer(hdi~years_from_2000+x1.6+x3.2+x5.1+x6.4+x7.3+(1+x1.6+x3.2+x5.1+x6.4+x7.3|region)+(1+years_from_2000||country),data=selected_columns)
```

The coefficients from the fixed effects are as follows:

```{r}
# Determine Significance of Coefficients
fixed_effects = as.data.frame(summary(time_mixed_effects)$coefficients)
adjusted_p_values = p.adjust(fixed_effects[, "Pr(>|t|)"], method = "bonferroni")

fixed_effects$`Bonferroni_Pr(>|t|)` = adjusted_p_values
fixed_effects
```
Although overly conservative, We have included Bonferroni adjused p-values, since multiple hypothesis tests were performed. The only fixed variable that was significant (both unadjusted and adjusted) in this model was the `years_from_2000` variable. The other variables were very far from significance. 

```{r}
# Importance of Considering Region
time_mixed_effects_noRegion = lmer(hdi~x1.6+x3.2+x5.1+x6.4+x7.3+(1+years_from_2000||country),data=selected_columns)
anova(time_mixed_effects,time_mixed_effects_noRegion)

# Importance of Considering Year
time_mixed_effects_noYear = lmer(hdi~x1.6+x3.2+x5.1+x6.4+x7.3+(1+x1.6+x3.2+x5.1+x6.4+x7.3|region),data=selected_columns)
anova(time_mixed_effects,time_mixed_effects_noYear)
```

We also analyzed the importance of including random effects of the variables `x1.6`, `x3.2`, `x5.1`, `x6.4`, and `x7.3` based on region, and the random effects of the time variable based on country using an extra-sum-of-squared F test. These F-tests

```{r}
# Make Plots
country_coef = as.data.frame(coef(time_mixed_effects)$country)
country_coef = merge(country_coef, selected_columns[,c("country", "region")], by.x=c("row.names"), by.y=c("country"))

ggplot(data = country_coef, aes(x=`(Intercept)`, y = years_from_2000)) + geom_point(aes(color=region))

residuals = augment(time_mixed_effects)
ggplot(data = residuals, aes(x=.fitted, y = .resid)) +
  geom_point(aes(color = region)) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  geom_hline(yintercept = 0) +
  labs(y="Residuals", x="FItted HDI")
```




















```{r}
# Old wrong model
time_mixed_effects = lmer(hdi~years_from_2000+x1.6+x3.2+x5.1+x6.4+x7.3+(1+years_from_2000|region),data=selected_columns)


residuals = augment(time_mixed_effects)
ggplot(data = residuals, aes(x=.fitted, y = .resid)) +
  geom_point(aes(color = region)) +
  theme(axis.text.x=element_blank(), axis.ticks.x=element_blank()) +
  geom_hline(yintercept = 0) +
  labs(y="Residuals", x="Fitted HDI")
```